---
title: "Final Data Project"
author: "DataTrail Team"
output:
  html_notebook:
    df_print: paged
---

# Share Results

## Final Project

Welcome to your final data storytelling project! This R Notebook template will guide you step-by-step as you build a compelling, reproducible, data-driven narrative.

Use the prompts in each section to shape your story, and share your analysis using this template.

## Objectives for this project

To complete this project, there are a few requirements you’ll need to fulfill. Remember that you’re not on your own for this project! Data science is best done as a community. If you get stuck or want to talk through your approach, don’t hesitate to ask questions. If you're in an in-person instructional setting, reach out to your instructors. You can also connect with others in your cohort, post on the community slack, or explore the Posit Community forums, where people of all experience levels share ideas, troubleshoot, and support each other.

This project should be completely uploaded and up to date on GitHub. Follow the steps in [Pushing and Pulling Changes](https://datatrail-jhu.github.io/DataTrail/pushing-and-pulling-changes.html) chapter for how to git add, commit, and push the changes you have done.

1.  At the very top of this file, put your own name in the `author:` place. Currently it says `"DataTrail Team"`, replace it with your name in the quotes. Also be sure to change the `title:` where it currently says `"Final Data Project"`.

2.  Follow good organization principles, you should at least have 2 folders: a `results` folder and a `data` folder.

3.  You should also have a README

4.  Make plots of results that you save to files.

5.  Write up your final observations in regards to your original question. Note that some data science projects end with "This isn't what I thought it would be" or "that's strange" or "I think this is leading into another question I would need to investigate". Whatever your observations may be, write them up in this R Markdown.

6.  When you feel your analysis is ready for review, send your instructor the GitHub link to your project so they can review it.

7.  Pat yourself on the back for all this hard work! You are well on your way to becoming a data scientist!

## Project Planning

This section is meant to help you define your research question and why it matters, similar to how we have done in the previous projects.

-   What is the research question you're exploring?

-   What have others explored about this topic?

-   Why is this topic interesting or important?

-   What do you hope to find or demonstrate?

*Note: If you need some inspiration on Telling Stories through data visualizations? A helpful resource is this [VOX article](http://www.vox.com/a/teens) using the YRBS data.*

## Data Sources

For this project, you have the freedom to choose your own dataset! Find something that sparks your curiosity and helps you dive into the question you want to explore. It’s your data adventure!

*Hint: You may refer back to our [Finding Data chapter](https://datatrail-jhu.github.io/DataTrail/finding-data.html) for more info on finding data.*

Some options for places to find data are:

-   [Kaggle](https://datatrail-jhu.github.io/DataTrail/finding-data.html#kaggle)
-   [FiveThirtyEight](https://datatrail-jhu.github.io/DataTrail/finding-data.html#fivethirtyeight-data)
-   You can also see datasets you already have in R by running this command: `ls("package:datasets")`.

You are not limited to these options for finding your data.

Describe the dataset(s) you are using. Include:

-   Dataset name, source and link (if public)

    -   Describe where you found your data and provide the link if applicable.

    -   Describe how the data was originally created. If your dataset is built into R (like one from the `datasets` package), you can learn more about it by using the help function—for example: `?datasets::AirPassengers`. If your dataset is from another source (like a file or online database), write a short summary explaining where the data came from, how it was collected, and any other useful context.

-   A brief description on what the data contains

## Goal

\<Write here what the goal of this analysis is. What question are we trying to answer?\>

## Setting up packages

Load the packages you will need for this analysis.

```{r}
## you can add more, or change...these are suggestions
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Setting up directories

Set up the directories you will need.

```{r}
if (!dir.exists("data")) {
  dir.create("data")
}
if (!dir.exists("results")) {
  dir.create("results")
}
```

## Getting the data

```{r}
# Read in your dataset(s)
```

## Exploring the data

It's important to get an initial sense of what your dataset looks like.

-   Look at the column names

-   Preview the first few rows

-   Summarize key statistics to get a sense of the data

```{r}

```

## Cleaning the data

Document any cleaning steps here.

-   Should you filter any rows?
-   Do you need to create new columns?
-   Are there any missing values?

*Reminder: The [dplyr](https://dplyr.tidyverse.org/) package was created for data wrangling, which is another way of saying data manipulation and cleaning. The six most commonly used data manipulation functions in this package are: `arrange`, `select`, `filter`, `mutate`, `summarize` and `group_by`.*

```{r}

```

## Plotting the data

Use this section to analyze your data and create plots that support your research question.

Reminder: [ggplot2](https://ggplot2.tidyverse.org/) is one of the most widely used packages for data visualization in R. It is one of the most common packages for creating plots.

```{r}
#ggplot(aes(VARIABLE)) +
  #geom_point()
```

## Calculating the stats

Use this section to perform statistical analyses and create visualizations based on those analyses. This includes exploring relationships between variables, for example, generating correlation plots or other relevant charts. Be sure to interpret any patterns or trends you observe and explain what they might suggest about the data.

```{r}

```

## Conclusion

Write up your thoughts about this data science project here and answer the following questions:

-   What did you find out in regards to your original question?
-   What exceptions or caveats do you have in regards to your analysis you did?
-   What did you find most challenging about this project?
-   What follow up questions do you have?

## Printing out session info

It’s good practice to include your session info at the end of your R Markdown notebook. This provides a snapshot of your R environment. It includes the version of R, the platform you’re running on, and the packages and libraries (including version numbers) that were loaded when the notebook was run.

This is helpful for:

-   Reproducibility (making sure others can run your code the same way)

-   Understanding what software versions were used

```{r}
sessionInfo()
```
