---
title: "Stats Project"
author: "DataTrail Team"
output:
  html_notebook:
    df_print: paged
---

# Stats

## Project: Country's literacy rates and life expectancy

In this project, you will utilize the same country datasets from project 3 to explore the relationship between literacy rates and life expectancy.

The dataset includes information from multiple global sources, including literacy rates and life expectancy data for both males and females. You will load, clean, and plot the data using tools from the tidyverse. Once the data is in good shape, you will use the correlation statistic to explore the relationships of numerous literacy and life expectancy variables.

## Objectives for this project

To complete this project you’ll need to do a few things within this file. As you go through the notebook, you will have further instruction on how to complete these objectives.

1.  At the very top of this file, put your own name in the `author:` place. Currently it says `"DataTrail Team"`, replace it with your name in the quotes.

2.  Go through this notebook, reading along.

3.  Fill in empty or incomplete code chunks when prompted to do so.

4.  Run each code chunk as you come across it by clicking the tiny green triangles at the right of each chunk. You should see the code chunks print out various output when you do this.

5.  In the Conclusions section at the end, write thoughtful responses to each of the questions provided.

6.  When done with this project notebook save the file by going to File \> Save in the RStudio menu. This will update the output `.nb.html` file to reflect your changes.

7.  Open up the resulting `countries_stats.nb.html` file and click “View in Web Browser”. Check to make sure everything looks good and your changes are showing up as expected.

8.  Upload your `Rmd` and your `nb.html` to your assignment folder (this is something that will be dependent on what your instructors have told you – or if you are taking this on your own, just collect these projects in one place, preferably a Google Drive).

## Goal

The goal of this analysis is:

\<Write here what the goal of this analysis is. What question are you trying to answer?\>

How do different country-level factors like population density, GDP, life expectancy, fertility rate, etc. relate to each other across the world?

## R Packages

For a refresher of what an R package is See [Data Trail Section 22.0.1 What is an R package?](https://datatrail-jhu.github.io/DataTrail/r-packages.html) or the Set up Packages section in the previous project notebook

These are the R packages you will be using:

-   [tidyverse](https://www.tidyverse.org/packages/)

-   [readr](https://readr.tidyverse.org/)

-   [dplyr](https://dplyr.tidyverse.org)

-   [ggplot2](https://ggplot2.tidyverse.org/)

-   [tidyr](https://tidyr.tidyverse.org/index.html)

-   [corrplot](https://rpubs.com/melike/corrplot)

-   [knitr](https://www.overleaf.com/learn/latex/Knitr)

## Setting up packages

Before using any package in R, first make sure it’s installed and ready to go! Then load the packages using the **`library()`** function including the name of the package in the parenthesis:

*Note: You can add packages as needed or not use/remove the ones not needed as you see fit.*

```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(corrplot)
library(knitr)
```

## Setting up directories

Here, you are making sure there’s a folder called `"data"` in your project. If it doesn’t already exist, it will be created it using the code below:

```{r}
if (!dir.exists("data")) {
  dir.create("data")
}
if (!dir.exists("plots")) {
  dir.create("plots")
}
```

## **Getting the data**

In a previous project in `03_Cleaning_Data`, we cleaned a dataset about countries and saved this to an RDS file. Read in this file from its location in the `03_Cleaning_Data` folder.

Hint: you will need to use a `..` to back out of this current folder and reference the parent folder of `05_Getting_Stats`.

```{r, solution = TRUE}
# Read in the dataset for countries_df
### Your code here
```

## Exploring the data

Use functions we have discussed previously to see remind yourself of what variables are in `countries_df`.

Hint: You can revisit chapter 18 section [18.1.1 Subsetting data frames](https://datatrail-jhu.github.io/DataTrail/data-frames-1.html) and [19.0.4 Are there named features of this object?](https://datatrail-jhu.github.io/DataTrail/basic-commands-in-r.html)

Write your code in the code chunk below:

```{r, solution = TRUE}
### Your code here
```

### Cleaning the data

You need to simplify the dataset by selecting columns that are most relevant to your analysis. This will make the dataset easier to work with and help you focus on exploring the relationships between key country-level indicators (e.g population, economy, health, and development). This will help with understanding global patterns and trends.

You will also rename some of the longer column names to shorter ones so that you can more easily manage these variables later.

```{r}
countries_df <- countries_df %>%
  select(country,
         pop = population_in_thousands_2017,
         pop_density = population_density_per_km2_2017,
         sex_ratio = sex_ratio_m_per_100_f_2017,
         gdp = gdp_per_capita_current_us,
         labour_force_participation_female_male_pop_percent,
         urban_perc = urban_population_percent_of_total_population,
         growth_rate = population_growth_rate_average_annual_percent,
         fertility = fertility_rate_total_live_births_per_woman,
         life_exp_female = life_expectancy_at_birth_females,
         life_exp_male =life_expectancy_at_birth_males,
         infant_mortality = infant_mortality_rate_per_1000_live_births,
         seats_held_women = seats_held_by_women_in_national_parliaments_percent,
         pop_using_improved_drinking_water_urban_rural_percent)
```

#### [**Breaking it down:**]{.underline}

-   `countries_df <- countries_df %>%`: This tells R to take the existing `countries_df` dataset and update it with a new version created by the steps that follow.

-   `select(...)`: This function keeps only the columns we list inside the parentheses and drops all the others.

-   `country`: Keeps the column called `country` as-is.

-   Inside `select()`, the syntax `new_name = old_name` tells R to “Keep `old_name`, but call it `new_name` from now on.”

    -   `pop = population_in_thousands_2017`: Keeps the `population_in_thousands_2017` column, but renames it to `pop`.

    -   `pop_density = population_density_per_km2_2017`: Keeps the population density column and renames it to `pop_density`.

    -   `sex_ratio = sex_ratio_m_per_100_f_2017`: Renames the column showing number of males per 100 females to `sex_ratio`.

    -   `gdp = gdp_per_capita_current_us`:Renames GDP per capita to `gdp`.

    -   `urban_perc = urban_population_percent_of_total_population`: Renames the column showing % of people in urban areas to `urban_perc`.

    -   `growth_rate = population_growth_rate_average_annual_percent`: Renames annual population growth rate to `growth_rate`.

    -   `fertility = fertility_rate_total_live_births_per_woman`: Renames fertility rate column to `fertility`.

    -   `life_exp_female = life_expectancy_at_birth_females`: Renames female life expectancy to `life_exp_female`.

    -   `life_exp_male = life_expectancy_at_birth_males`: Renames male life expectancy to `life_exp_male`.

    -   `infant_mortality = infant_mortality_rate_per_1000_live_births`: Renames infant mortality rate to `infant_mortality`.

-   `labour_force_participation_female_male_pop_percent`: Keeps this column as-is (no renaming).

-   `seats_held_women = seats_held_by_women_in_national_parliaments_percent`: Renames column showing % of women in national parliament to `seats_held_women`.

-   `pop_using_improved_drinking_water_urban_rural_percent`: Keeps this column as-is (no renaming).

There are some variables here that need to be split up into two columns much like we did before in `03_Cleaning_Data`.

Hint: *You can revisit chapter 32 section [32.0.9 Separating Columns](https://datatrail-jhu.github.io/DataTrail/how-to-tidy-data.html).*

`labour_force_participation_female_male_pop_percent` `pop_using_improved_drinking_water_urban_rural_percent`

Write your code in the code chunk below:

```{r, solution = TRUE}
### Your code here
```

Now all of the variables in this dataset, except for `country`, should be numeric but that might not be the case. Try using `as.numeric` within `mutate` on all of the variables except `country`.

*Hint: In an earlier project DT 02 you used `across` to `mutate` multiple columns (variables) you wanted to change at once. You can learn more about mutating multiple columns [here](https://dplyr.tidyverse.org/reference/across.html), where they explain how to use `across`. You will need an opperator that specifies exclude the `country` column.*

*Note: A warning may appear that happens when R tries to convert a value to a number (`as.numeric`) but can't (for example, if a column contains text or special characters). R will then turn those values into `NA` (missing values). This is just a warning and doesn't stop your code from running.*

Write your code in the code chunk below:

```{r, solution = TRUE}
### Your code here
```

Save this version of your data to an RDS file in this project.

\
*Hint: revisit chapter 24 section [24.0.5 Writing CSV files into Posit Cloud](https://datatrail-jhu.github.io/DataTrail/dataset-file-types.html) and [24.1.0.2 Serialized R objects (.rds)](https://datatrail-jhu.github.io/DataTrail/dataset-file-types.html)*

```{r, solution = TRUE}
### Your code here
```

## Plotting the data

Let's explore the kinds of distributions these variables have.

Use ggplot2::geom_density() to do this for at least 5 variables of your choice in `countries_df`.

To better understand the patterns and ranges in this dataset, you will explore the distribution of several key variables using `geom_density()` from the ggplot2 package.

Create density plots for at least 5 variables of interest from `countries_df`. Write your code in the code chunks below:

```{r, solution = TRUE}
### Your code here
```

```{r, solution = TRUE}
### Your code here
```

```{r, solution = TRUE}
### Your code here
```

```{r, solution = TRUE}
### Your code here
```

```{r, solution = TRUE}
### Your code here
```

```{r, solution = TRUE}
### Your code here
```

## Calculating the stats

Now you are ready to explore how the variables relate to each other. You can use correlation analysis, which helps with seeing the relationship with numeric variables if they tend to increase or decrease together. You will use the `cor()` function to calculate correlation coefficients (values between -1 and 1) that tell you how strong the relationship is between two variables. If you remember from earlier projects: a value closer to 1 means a strong positive relationship while a value closer to -1 means a strong negative relationship.

Use the `cor()` function to create a correlation matrix of all the numeric variables in `countries_df`.

*Hint: chapter 53 has section [53.5.3 Correlation tests](https://datatrail-jhu.github.io/DataTrail/in-practice-using-stats.html) but remember that if you need help with how the `cor()` function works you can type `?cor` into your Console to pull up the help page. This will show you the function's arguments, examples, and how to use it.*

*Hint: If you’re getting NAs as a result in the table, check whether your data contains missing values. There’s an argument you can add to the function to handle this by excluding incomplete rows.*

Write your code in the code chunk below:

```{r, solution = TRUE}
### Your code here
```

Now, format the correlation matrix into a readable table.

*Hint: You can revisit chapter 46 section [46.0.3 Improving the table output](https://datatrail-jhu.github.io/DataTrail/tables-in-r.html).*

Write your code in the code chunk below:

```{r, solution = TRUE}
### Your code here
```

Let’s create a visual version of the correlation matrix using `corrplot()` from the corrplot package. This helps us quickly spot strong relationships or interesting groupings between variables.

Use the `corrplot::corrplot()` function to make a correlation matrix.

Hint: You can see a quick overview on the [corrplot() function](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html).

Write your code in the code chunk below:

```{r, solution = TRUE}
### Your code here
```

What do you think about the interaction and groupings of these variables? Is there anything surprising?

Save your plot in the plots folder. Write your code in the code chunk below:

```{r, solution = TRUE}
### Your code here
```

### Further exploration

Now that you’ve seen how variables in the dataset relate to one another using a correlation matrix, let’s take it a step further.

Choose two numeric variables from `countries_df` that seem to have a meaningful relationship based on the correlation plot. Visualize the relationship with a plot of your choice.

Write your code in the code chunk below:

```{r}
# Make a plot of your choice here
```

Choose two numeric variables from `countries_df` that seem to have a meaningful relationship based on the correlation plot. Run a statistical test that you think may work best to explore how strongly the two variables are related.

Hint: you may want to revisit chapter 53 section [53.5 How do I know what test to use?](https://datatrail-jhu.github.io/DataTrail/in-practice-using-stats.html)

Write your code in the code chunk below:

```{r}
# Run a statistical test of your choice here
```

## Conclusion

Write up your thoughts about this data science project here and answer the following questions:

-   What have you learned about the relationship of the two variables you chose?

    I explored the relationship between GDP per capita and life expectancy for females. The correlation matrix and visualization both showed a clear positive relationship with countries with higher GDP per capita tending to have higher female life expectancy. This suggests that economic resources may be linked to better healthcare, nutrition, and living conditions that contribute to longer lives.

-   Was anything surprising or confusing?

    It was surprising to see that not all variables with strong social relevance (like percentage of women in parliament or access to improved drinking water) had strong correlations with life expectancy or GDP. It’s a good reminder that meaningful relationships don’t always show up clearly in numeric comparisons.

-   What alternative hypothetical explanations are there for the observed relationships you are seeing? (Remember correlation is not causation, what hidden factors could drive both variables?)

    The correlation between GDP and life expectancy doesn’t necessarily mean one causes the other. It’s possible that a third variable (e.g education level, health infrastructure, or government stability) is influencing both. Correlation can highlight patterns, but it doesn’t explain why those patterns exist.

-   In a hypothetical future study, what questions would you want to explore and what kind of data would you want to use to do this exploration?

    In a hypothetical future study, it would be interesting to explore whether gender equity (like labour force participation or seats held by women) influences health outcomes or development indicators. To do this, perhaps a more detailed time-series of data would be helpful to look at trends over time and maybe additional variables like education levels, healthcare access, or government spending on public health and education.

## Printing out session info

It’s good practice to include your session info at the end of your R Markdown notebook. This provides a snapshot of your R environment. It includes the version of R, the platform you’re running on, and the packages and libraries (including version numbers) that were loaded when the notebook was run.

This is helpful for:

-   Reproducibility (making sure others can run your code the same way)

-   Understanding what software versions were used

```{r}
sessionInfo()
```
